---
title: "How to use GeoPressureR: an example"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to use GeoPressureR: an example}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r, include = FALSE}
knitr::opts_chunk$set(
  warning = FALSE, 
  message = FALSE,
  collapse = TRUE,
  comment = "#>"
)
# knitr::knit("vignettes/basic_example.Rmd.orig", output = "vignettes/basic_example.Rmd")
```
Using a basic example, we walk through the main steps of the methodology to demonstrate how it can be used. 

## Preparing the data

To start, install the GeoPressureR package from Github using the following line: 
```{r, eval = FALSE}
devtools::install_github("rafnuss/GeoPressureR")
```

We will be using the following libraries: 
```{r setup}
library(GeoPressureR)
library(raster)
library(leaflet)
library(ggplot2)
library(plotly)
library(RColorBrewer)
```

### Reading geolocator data
In this example, we use data captured on a Great Reed Warbler (18LX). Below, we read the geolocator data and define the start and end time (crop_start and crop_end).

```{r}
pam_data = read_pam(pathname = system.file("extdata", "/", package = "GeoPressureR") )
crop_start = as.POSIXct("2017-06-20","%Y-%m-%d", tz="UTC")
crop_end = as.POSIXct("2018-05-02","%Y-%m-%d", tz="UTC")
pam_data = crop_pam(pam_data, crop_start, crop_end)
```

### Automatic classification of activity

We use a simple k-mean clustering to define periods of relatively high activity, and then classify high activities lasting more than 30 minutes as migratory activities. See more possible classifications described in the [PALMr manual](https://kiranlda.github.io/PAMLrManual/index.html).

```{r}
pam_data = classify_pam(pam_data, min_duration = 30)
```

### Editing on TRAINSET

To ensure the high level of precision needed for the pressure match, we must manually edit the activity classification and  the pressure timeseries to be matched. We suggest doing this with TRAINSET. A separate vignette dedicated to this exercise, including best practices and a sample code to get started, is available here [labeling track vignette](labeling_tracks.html).

Use `trainset_write()` to export the automatically generated classifications in a `*.csv` file, which can be opened in [TRAINSET](https://trainset.geocene.com/).

```{r eval=F}
trainset_write(pam_data, pathname=system.file("extdata", "/", package = "GeoPressureR"))
# browseURL("https://trainset.geocene.com/")
```


![Printscreen of the manual classification in Trainset. See [labeling track vignette](labeling_tracks.html) for more information](labeling_tracks_4.png){width=100%}

When you have finished the manual editing, export the new `*-labeled.csv` file. Make sure to keep these classified files in your project folder (e.g. under `/data/`).

To re-edit an existing labeled file, you can simply re-open the file on TRAINSET and read this file directly with `trainset_read()`.

```{r}
pam_data = trainset_read(pam_data, pathname=system.file("extdata", "/", package = "GeoPressureR"))
```

### Identifying stationary periods

Based on the activity labeling, `sta_pam()` creates a table of stationary periods, illustrated below.

```{r}
pam_data = sta_pam(pam_data)
knitr::kable(head(pam_data$sta))
```

We can visualize the pressure measurements for each grouped stationary period (symbolized by the same color).

```{r}
p <- subset(pam_data$pressure, sta_id != 0) %>% 
  ggplot() +
  geom_line(aes(x=date,y=obs,col=as.factor(sta_id))) + 
  theme_bw() +
  scale_y_continuous(name="Pressure(hPa)") +
  scale_colour_manual(values=rep(RColorBrewer::brewer.pal(9,"Set1"),times=4))
  #scale_colour_brewer(type="qualitative", palette = 'Set1')

ggplotly(p, dynamicTicks = T)
```

## Request map of pressure
We can now clearly visualize the unique pressure timeseries for each stationary period, and are ready to match each one with atmospheric pressure data (ERA5). We use Google Earth Engine to overcome the challenge of computing mismatch on such a large dataset.

Initially, it might be easier to query only stationary periods of a certain duration.
```{r}
sta_id_keep = pam_data$sta$sta_id[difftime(pam_data$sta$end,pam_data$sta$start, units = "hours")>12]
pam_data$pressure$sta_id[!(pam_data$pressure$sta_id %in% sta_id_keep)] = NA
```

We can query the data. To save time in this vignette I load this data that I've saved earlier. But the code is given below
```{r}
data("prob_map_list", package = "GeoPressureR")
data("ts_list", package = "GeoPressureR")
```

In this part, we query the pressure maps from the [GeoPressure API](https://github.com/Rafnuss/GeoPressureServer/).
```{r, eval=F}
extent = c(-16,20,0,50) # coordinates of the map to request (W,E,S,N) 
scale = 10 # request on a 0.1° grid to make the code faster
max_sample = 250
margin = 30 # roughly equivalent to 3hPa
raster_list = geopressure_map(pam_data$pressure, extent = extent, scale = scale, max_sample = max_sample, margin = margin)
```

Compute probability map
```{r, eval = F}
s = 0.4 # standard deviation of pressure
thr = 0.9 # threashold of value acceptable 2
prob_map_list = geopressure_prob_map(raster_list, s=s, thr=thr)
```

### Display
Comparing maps of mismatch, threadhold and probability
```{r, eval = F}
i_r = 1;

leaflet() %>% addTiles() %>%
  addRasterImage(prob_map_list[[i_r]], opacity = 0.8, group="Probability") %>%
  addRasterImage(raster_list[[i_r]][[1]], opacity = 0.8, group="Mismatch") %>%
  addRasterImage(raster_list[[i_r]][[2]], opacity = 0.8, group="Threashold") %>%
  # addLegend(pal = pal, values = values(v[[i_s]][[3]]), title = "Probability") %>% 
  addLayersControl(
    overlayGroups = c("Probability","Mismatch","Threashold"),
    options = layersControlOptions(collapsed = FALSE)
  ) %>% hideGroup(c("Mismatch","Threashold"))
```
```{r, eval = F}

li_s=list()
l = leaflet() %>% addTiles() 
for (i_r in 1:length(prob_map_list)){
  i_s = metadata(prob_map_list[[i_r]])$sta_id
  
  info = pam_data$sta[pam_data$sta$sta_id==i_s,]
  info_str = paste0(i_s," | ",info$start,"->",info$end)
  li_s <- append(li_s, info_str)
  l = l %>% addRasterImage(prob_map_list[[i_r]], opacity = 0.8, group=info_str) 
}

l %>% 
  addLayersControl(
    overlayGroups = li_s,
    options = layersControlOptions(collapsed = FALSE)
  ) %>% hideGroup(tail(li_s,length(li_s)-1))
```


## Check pressure match
```{r}
i_r=1

i_s = metadata(prob_map_list[[i_r]])$sta_id

# find the max value of probability
tmp = as.data.frame(prob_map_list[[i_r]],xy=T)
lon = tmp$x[which.max(tmp[,3])]
lat = tmp$y[which.max(tmp[,3])]

# filter pressure for the stationary period
# include flight period before and after
id = pam_data$pressure$sta_id==i_s & !is.na(pam_data$pressure$sta_id)
pressure = list(
  obs = pam_data$pressure$obs[id],
  date = pam_data$pressure$date[id],
  class = pam_data$pressure$class[id]
)
```

Query the pressure at this location

```{r, eval=F}
ts_list[[i_r]]  = geopressureTs(lon, lat, pressure = pressure)
```

Visualize
```{r}
ts_list[[i_r]]$pressure0 = ts_list[[i_r]]$pressure - mean(ts_list[[i_r]]$pressure) + mean(pressure$obs)

# rbind(as.data.frame(pressure),as.data.frame(ts_list))

p <- ggplot() +
  geom_point(data=as.data.frame(pressure), aes(x=date,y=obs,col="Geolocator")) +
  geom_line(data=as.data.frame(ts_list[[i_r]]), aes(x=date,y=pressure0,col="ERA5 Reanalyis")) + 
  theme_bw()

ggplotly(p) #%>% layout(legend = list(orientation = "h", x = -0.5))
```

