---
title: "Preparing data for trajectory modelling"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Preparing data for trajectory modelling}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, message=F}
library(GeoPressureR)
library(raster)
library(leaflet)
```

## Load data
```{r}
data("pressure_prob", package = "GeoPressureR")
pressure_prob = pressure_prob
data("light_prob", package = "GeoPressureR")
pam_data = pam_read(pathname = system.file("extdata", package = "GeoPressureR"),
                    crop_start = "2017-06-20", crop_end = "2018-05-02")
pam_data = trainset_read(pam_data, pathname=system.file("extdata", package = "GeoPressureR"))
pam_data = pam_sta(pam_data)
```

## Align and pressure and light
Because pressure might not be computed on all stationary period, we need to filter the map of light and compute the equivalent flight duration

```{r}
sta_pres = unlist(lapply(pressure_prob, function(x) raster::metadata(x)$sta_id))
sta_light = unlist(lapply(light_prob, function(x) raster::metadata(x)$sta_id))

# filter light map
light_prob = light_prob[sta_light %in% sta_pres]

# compute flight  duration
grp_id = seq_len(nrow(pam_data$sta)) %in% sta_pres
sta = pam_data$sta[grp_id,]
sta$all_next_flight_duration =  aggregate(pam_data$sta$next_flight_duration, by=list(Category=cumsum(grp_id)), sum )$x
```

Compute the static probability with the product of light and pressure probability map. We also add the flight duration in the metadata

```{r}
static_prob = mapply(function(light,pressure,next_flight_duration){
  static_prob = light * pressure
  tmp <- values(static_prob)
  tmp[is.na(tmp)] <- 0
  values(static_prob) <- tmp
  # keep metadata
  metadata(static_prob)  <- metadata(pressure)
  metadata(static_prob)$next_flight_duration  <- next_flight_duration
  # return
  static_prob
}, light_prob, pressure_prob,as.numeric(sta$all_next_flight_duration, units="hours"))
```

## Downscaling

It is possible to downscale the map with the following code. But we don't use it here.

```{r, eval=F}
static_prob = lapply(static_prob, function(raster){
  raster_ds  <- aggregate(raster , fact=1, fun = max, na.rm=T, expand=T)
  # keep metadata
  metadata(raster_ds)  <- metadata(raster)
  # return
  raster_ds
})
```

## Add equipement and retrival information

Overwrite probability at the equipment and retrival site

```{r}
lon_calib <- 17.05
lat_calib <- 48.9

lat = seq(raster::ymax(static_prob[[1]]), raster::ymin(static_prob[[1]]), length.out=nrow(static_prob[[1]])+1)
lat = lat[1:length(lat)-1]+diff(lat[1:2])/2
lon = seq(raster::xmin(static_prob[[1]]), raster::xmax(static_prob[[1]]), length.out=ncol(static_prob[[1]])+1)
lon = lon[1:length(lon)-1]+diff(lon[1:2])/2

lon_calib_id = which.min(abs(lon_calib-lon))
lat_calib_id = which.min(abs(lat_calib-lat))

tmp = as.matrix(static_prob[[1]])
tmp[,] <- 0
tmp[lat_calib_id,lon_calib_id] <- 1
values(static_prob[[1]]) <- tmp

tmp = as.matrix(static_prob[[length(static_prob)]])
tmp[,] <- 0
tmp[lat_calib_id,lon_calib_id] <- 1
values(static_prob[[length(static_prob)]]) <- tmp
```


## Visualization

```{r, warning=F}
li_s=list()
l = leaflet() %>% addTiles() 
for (i_r in 1:length(static_prob)){
  i_s = metadata(static_prob[[i_r]])$sta_id
  info = metadata(static_prob[[i_r]])$extend_sample
  info_str = paste0(i_s," | ",info[1],"->",info[2])
  li_s <- append(li_s, info_str)
  l = l %>% addRasterImage(static_prob[[i_r]], opacity = 0.8, group=info_str) 
}
l %>% 
  addLayersControl(
    overlayGroups = li_s,
    options = layersControlOptions(collapsed = FALSE)
  ) %>% hideGroup(tail(li_s,length(li_s)-1))
```

```{r, animation.hook="gifski", eval=F}
library(gifski)
raster::animate(brick(static_prob),n=1)
```

```{r, eval=F}
library(ggplot2)
library(gganimate)
dfb = do.call("rbind",lapply(static_prob,function(x){
  mt = metadata(x)
  df = as.data.frame(x,xy=T)
  df$sta_id=mt$sta_id
  df$extend_sample = mt$extend_sample
  df$next_flight_duration = mt$next_flight_duration
  df
}))

dfb %>% 
  ggplot() +
  geom_raster(aes(x=x, y=y, fill=layer)) +
  transition_time(sta_id)
```

## Check compatibility

First check that at a probability exist at each stationary period.
```{r}
for (i_s in seq_len(length(static_prob))){
  if (sum(values(static_prob[[i_s]]), na.rm=T)==0){
    stop(paste('Probability map of stationary period', i_s, 'is null. Check part 1 process (light and pressure)', sep=' '))
  }
}
```

Then check that there are always at least one transition possible from one stationary period to the next

```{r}
for (i_s in seq_len(length(static_prob)-1)){
  cur = as.matrix(static_prob[[i_s]])>0
  nex = as.matrix(static_prob[[i_s+1]])>0
  
  flight_duration = metadata(static_prob[[i_s]])$next_flight_duration # hours
  resolution = mean(res(static_prob[[1]]))*111 # assuming 1°= 111km
  thr_gs = 150 # Assuming a max groundspeed of 150km/h
  
  # Check possible position at next stationary period 
  possible_next = (EBImage::distmap(!cur) * resolution /flight_duration)<thr_gs
  
  if (sum(possible_next & nex)==0){
    stop(paste('There are no possible transition from stationary period', i_s, 'to',i_s+1,'. Check part 1 process (light and pressure)', sep=' '))
  }
}
```


## Export data
We can export the data generated.

```{r, eval=F}
usethis::use_data(static_prob, overwrite=T)
```
